---
title: "AIT Final"
author: "Devin Rohler"
date: "11/5/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#source file for aesthetics 
setwd("/Users/Devin/Desktop/School Work/Graduate/Fall 2021/AIT 580/Project")
```

```{r}
# Read in and display the data file 
crime <- read.csv('c:/Users/Devin/Desktop/School Work/Graduate/Fall 2021/AIT 580/Project/NYC_CRIME_2014.csv', header=TRUE, sep = ',')
```

```{r}
#packages
library(tidyr)
library(tidyverse)
library(ggplot2)
library(stringr)
library(ggridges)
```



Data Manipulation for Processing and Exploratory Analysis
```{r}
#Rename
crime <- crime %>% rename(date = ï..DATE, time = TIME.OCC, area_num = AREA, area_name = AREA.NAME, sub_area_num = Rpt.Dist.No,
                 crime_code = Crm.Cd, crime_description = Crm.Cd.Desc, vic_age = Vict.Age, vic_sex = Vict.Sex, vic_descent = Vict.Descent, structure_code = Premis.Cd,
                 structure_description = Premis.Desc, weapon_code = Weapon.Used.Cd, weapon_description = Weapon.Desc)
```
```{r}
#string manipulation for date variable, cut off end, and split into 3 different variables 
#remove parenthesis
type <- str_split_fixed(crime$crime_description,'[()]',2)
type <- data.frame(type)
type <- subset(type, select = c(X1))
type <- type %>%  rename(crime_type = X1)

#remove parenthesis
structure <- str_split_fixed(crime$structure_description,'[()]',2)
structure <- data.frame(structure)
structure <- subset(structure, select = c(X1))
structure <- structure %>% rename(premis = X1)

#remove parenthesis 
weapon <- str_split_fixed(crime$weapon_description,'[()]',2)
weapon <- data.frame(weapon)
weapon <- subset(weapon, select = c(X1))
weapon <- weapon %>% rename(weapon_type = X1)

#remove parenthesis 
crime_split <- str_split_fixed(crime$date,' ',2)
crime_split <- data.frame(crime_split)

#remove parenthesis
crime_n <- str_split_fixed(crime_split$X1,'/',3)
crime_n <- data.frame(crime_n)
crime_n <- crime_n %>% rename(month = X1, day = X2, year = X3)

crime_1 <- subset(crime, select = -c(date))
#combine str manipulated variables
crime <- cbind(type,structure,weapon,crime_n,crime_1)
#drop codes, have type variable as well, code is not needed
crime <- subset(crime, select = -c(area_num,crime_description,structure_description,weapon_description))
```
```{r}
#recode variables
crime$vic_sex = factor(crime$vic_sex,
                   levels = c('M','F','X'),
                   labels = c("Male","Female","Unknown"))
crime$vic_descent = factor(crime$vic_descent,
                   levels = c('A','B','C','D','F','G','H','I','J'
                              ,'K','L','O','P','S','U','V','W','X','Z'),
                   labels = c('Other Asian','Black','Chinese','Cambodian',
                              'Filipino','Guamanian','Hispanic','American Inidan',
                              'Japanese','Korean','Laotian','Other','Pacific Islander',
                              'Samoan','Hawaiian','Vietnamese','White',
                              'Unknown','Asian Indian'))

crime$month <- factor(crime$month, levels = c("1","2","3","4","5","6","7","8","9","10",
                                              "11","12"),
                      labels = c("Jan","Fed","Mar","Apr",
                      "May","Jun","Jul","Aug","Sep","Oct","Nov","Dec"))
```



Exploratory Analysis 

```{r}
#Crime Count/Month
crime %>% group_by(month) %>% 
  summarise(n_crimes = n()) %>% 
  arrange(desc(n_crimes))
```
```{r, fig.height= 7, fig.width= 9}
#Crime Count Time Line
ggplot(crime,aes(x = month)) + 
  geom_bar(fill = '#CC6600', color =  'black') +
  labs(title = "Number of Crimes Per Month for L.A.",
       subtitle = "For the Year 2014",
       x = "Month",
       y = "# of Crimes")
```




```{r}
#crime types
crime %>% group_by(crime_type) %>% 
  summarise(n_crimes = n()) %>% 
  arrange(desc(n_crimes), n = 10)
```
```{r}
#crime weapon
crime %>% group_by(weapon_type) %>% 
  summarise(weapon_count = n()) %>% 
  arrange(desc(weapon_count))
```
```{r}
#crime premis
crime %>% group_by(premis) %>% 
  summarise(location_count = n()) %>% 
  arrange(desc(location_count))
```
```{r}
#crime location
crime %>% group_by(area_name) %>% 
  summarise(area_count = n()) %>% 
  arrange(desc(area_count))
```
```{r}
#victum descent
crime %>% group_by(vic_descent) %>% 
  summarise(vic_descent_count = n()) %>% 
  arrange(desc(vic_descent_count))
```
```{r}
#victum sex
crime %>% group_by(vic_sex) %>% 
  summarise(vic_sex_count = n()) %>% 
  arrange(desc(vic_sex_count))
```
```{r}
#Average Number of Crimes/Day in 2014
crime %>% group_by(year) %>%
  summarise(crime_count = n()/365)
```



Research Questions 
 1. What impact do major holidays have on the number of crimes in L.A.?
```{r}
#Fourth of July Crime Types
fourth <- filter(crime, month == 'Jul' & day == 4)
fourth %>% group_by(crime_type) %>%
  summarize(crime_4th_count = n()) %>% 
  arrange(desc(crime_4th_count))
```
```{r}
#Total on 4th
fourth %>% group_by(day) %>% 
  summarize(total_4th = n())
```


```{r}
#Thanksgiving Crime Types
thanks <- filter(crime, month == 'Nov' & day == 27)
thanks%>% group_by(crime_type) %>%
  summarize(crime_thanks_count = n()) %>% 
  arrange(desc(crime_thanks_count))
```
```{r}
#Total on Thanksgiving
thanks %>% group_by(day) %>% 
  summarize(total_thanks = n())
```


```{r}
#Christmas Crime Types
xmas <- filter(crime, month == 'Dec' & day == 25)
xmas %>% group_by(crime_type) %>%
  summarize(crime_xmas_count = n()) %>% 
  arrange(desc(crime_xmas_count))
```
```{r}
#Total on Christmas
xmas %>% group_by(day) %>% 
  summarize(total_xmas = n())
```


```{r}
#New Years Crime Types
new <- filter(crime, month == 'Jan' & day == 1)
new %>% group_by(crime_type) %>%
  summarize(crime_new_count = n()) %>% 
  arrange(desc(crime_new_count))
```
```{r}
#Total on New Years
new %>% group_by(day) %>% 
  summarize(total_new = n())
```



2. How does the percent of victums by gender and race change during these Holidays, in L.A.?

This questions will be analyzed by showing statistics of gender,race and both and then comparing it to the yearly average

#Fourth of July
```{r}
#Fourth, sex
fourth %>% group_by(vic_sex) %>%
  summarize(fourth_2_count =n()) %>%
  arrange(desc(fourth_2_count))
```
```{r}
#Fourth, race
fourth %>% group_by(vic_descent) %>%
  summarize(fourth_2_count =n()) %>%
  arrange(desc(fourth_2_count))
```
```{r}
#Fourth, both
fourth %>% group_by(vic_sex,vic_descent) %>%
  summarize(fourth_2_count =n()) %>%
  arrange(desc(fourth_2_count))
```
```{r}
#Fourth, both
fourth %>% group_by(vic_sex,vic_descent,crime_type) %>%
  summarize(fourth_2_count =n()) %>%
  arrange(desc(fourth_2_count))
```

```{r, fig.height= 7, fig.width= 9}
#Fourth Plot
ggplot(fourth,aes(x = vic_descent, fill = vic_sex)) + 
  geom_bar(color =  'black') +
  labs(title = "Number of Crimes Per Race & Gender for L.A.",
       subtitle = "For July 4th, 2014",
       x = "Victim Race",
       y = "# of Crimes") +
  guides(fill=guide_legend(title = "Victim Sex"))
```


#Thanksgiving
```{r}
#Thanksgiving, sex
thanks %>% group_by(vic_sex) %>%
  summarize(thanks_2_count =n()) %>%
  arrange(desc(thanks_2_count))
```
```{r}
#Thanksgiving, race
thanks %>% group_by(vic_descent) %>%
  summarize(thanks_2_count =n()) %>%
  arrange(desc(thanks_2_count))
```
```{r}
#Thanksgiving, both
thanks %>% group_by(vic_descent,vic_sex) %>%
  summarize(thanks_2_count =n()) %>%
  arrange(desc(thanks_2_count))
```
```{r}
#Thanksgiving, both
thanks %>% group_by(vic_descent,vic_sex,crime_type) %>%
  summarize(thanks_2_count =n()) %>%
  arrange(desc(thanks_2_count))
```
```{r, fig.height= 7, fig.width= 9}
#Thanksgiving Plot
ggplot(thanks,aes(x = vic_descent, fill = vic_sex)) + 
  geom_bar(color =  'black') +
  labs(title = "Number of Crimes Per Race & Gender for L.A.",
       subtitle = "For Thanksgiving 2014",
       x = "Victim Race",
       y = "# of Crimes") +
    guides(fill=guide_legend(title = "Victim Sex"))
```



#Christmas
```{r}
#Christmas, sex
xmas %>% group_by(vic_sex) %>%
  summarize(xmas_2_count =n()) %>%
  arrange(desc(xmas_2_count))
```
```{r}
#Christmas, race
xmas %>% group_by(vic_descent) %>%
  summarize(xmas_2_count =n()) %>%
  arrange(desc(xmas_2_count))
```
```{r}
#Christmas
xmas %>% group_by(vic_descent,vic_sex) %>%
  summarize(xmas_2_count =n()) %>%
  arrange(desc(xmas_2_count))

```
```{r}
#Christmas
xmas %>% group_by(vic_descent,vic_sex,crime_type) %>%
  summarize(xmas_2_count =n()) %>%
  arrange(desc(xmas_2_count))
```
```{r, fig.height= 7, fig.width= 9}
#Christmas Plot Plot
ggplot(xmas,aes(x = vic_descent, fill = vic_sex)) + 
  geom_bar(color =  'black') +
  labs(title = "Number of Crimes Per Race & Gender for L.A.",
       subtitle = "For Christmas 2014",
       x = "Victim Race",
       y = "# of Crimes") +
    guides(fill=guide_legend(title = "Victim Sex"))
```


#New Years
```{r}
#New Years, sex
new %>% group_by(vic_sex) %>%
  summarize(thanks_2_count =n()) %>%
  arrange(desc(thanks_2_count))
```
```{r}
#New Years, race
new %>% group_by(vic_descent) %>%
  summarize(thanks_2_count =n()) %>%
  arrange(desc(thanks_2_count))
```
```{r}
#New Years, both
new %>% group_by(vic_descent,vic_sex) %>%
  summarize(new_2_count =n()) %>%
  arrange(desc(new_2_count))
```
```{r}
#New Years, both
new %>% group_by(vic_descent,vic_sex,crime_type) %>%
  summarize(new_2_count =n()) %>%
  arrange(desc(new_2_count))
```
```{r, fig.height= 7, fig.width= 9}
#New years Plot
ggplot(new,aes(x = vic_descent, fill = vic_sex)) + 
  geom_bar(color =  'black') +
  labs(title = "Number of Crimes Per Race % Gender for L.A.",
       subtitle = "For New Years 2014",
       x = "Victim Race",
       y = "# of Crimes") + 
      guides(fill=guide_legend(title = "Victim Sex"))
```


#Yearly Average to compare
```{r}
#Yearly Average Crime Per Victim Sex
crime %>% group_by(vic_sex) %>%
  summarize(average_sex_count =n()/365) %>%
  arrange(desc(average_sex_count))
```
```{r}
#Yearly Average Crime Per Victim Race
crime %>% group_by(vic_descent) %>%
  summarize(average_race_count =n()/365) %>%
  arrange(desc(average_race_count))
```
```{r}
#Yearly Average Crime Per Victim sex and race combined
crime %>% group_by(vic_sex,vic_descent) %>%
  summarize(average_race_count =n()/365) %>%
  arrange(desc(average_race_count))
```




3. How do the crime rates change for areas of L.A. during these holidays?
```{r}
#Fourth
fourth %>% group_by(area_name) %>%
  summarize(fourth_3_count =n()) %>%
  arrange(desc(fourth_3_count))
```
```{r}
#Thanksgiving 
thanks %>% group_by(area_name) %>%
  summarize(thanks_3_count =n()) %>%
  arrange(desc(thanks_3_count))
```
```{r}
#Thanksgiving 
xmas %>% group_by(area_name) %>%
  summarize(xmas_3_count =n()) %>%
  arrange(desc(xmas_3_count))
```
```{r}
#New Years
new %>% group_by(area_name) %>%
  summarize(new_3_count =n()) %>%
  arrange(desc(new_3_count))
```


#Yearly Average for location
```{r}
#Average Number of Crimes/Location for LA, 2014
crime %>% group_by(area_name) %>% 
  summarise(area_count = n()/365) %>% 
  arrange(desc(area_count))
```

































