---
title: "STAT 515 Final"
author: "Devin Rohler"
date: "10/22/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#package
library(tidyr)
library(tidyverse)
library(ggplot2)
library(reshape2)
library(gridExtra)
library(corrplot)
library(car)
library(GGally)
library(randomForest)
library(rpart)
library(rpart.plot)
library(DMwR2)
library(ggcorrplot)
library(RColorBrewer)
library(useful)
```


1. "How do certain variables effect death?  Do some variables have a higher effect on death as an outcome than others?"

2. "Can a model be created to predict whether a patient will die or not based on the avaliable variables?"

```{r}
#Read in
heart <- read.table(file.choose(), sep = ",", header = TRUE)
```
```{r}
#source file for aesthetics 
source(file = "C:/Users/Devin/Desktop/School Work/Graduate/Fall 2021/STAT 515/Final/hw.r")
```
```{r}
#tidy
heart <- rename(heart, creatinine_ph = creatinine_phosphokinase, creatinine_se = serum_creatinine, death = DEATH_EVENT)
```
```{r}
#Recode
heart$sex = factor(heart$sex,
                   levels = c(0,1),
                   labels = c("Female","Male"))
heart$death = factor(heart$death,
                     levels = c(0,1),
                     labels = c('Alive','Dead'))
heart$anaemia = factor(heart$anaemia,
                     levels = c(0,1),
                     labels = c('False','True'))
heart$high_blood_pressure = factor(heart$high_blood_pressure,
                     levels = c(0,1),
                     labels = c('False','True'))
heart$diabetes = factor(heart$diabetes,
                     levels = c(0,1),
                     labels = c('False','True'))
heart$smoking = factor(heart$smoking,
                     levels = c(0,1),
                     labels = c('False','True'))
```


Dataset Preview and Stats
```{r}
heart %>% group_by(sex, death) %>% 
  summarise(n_patients = n(),
             min_age = min(age),
             avg_age = mean(age),
             max_age = max(age)) %>% 
    mutate(freq = n_patients / sum(n_patients))


#From this grouped dataframe we can get a better look into the dataset. This shows how many men and women make up the data set and alittle about thier ages
#The data includes a total of 299 patients, 105(women) and 194(male)
#Out of those 105 women, 34 died and out of the 194 men 62 died
```
```{r, fig.width=11, fig.height=8}
#Visulization #1
ggplot(data = heart,aes(x = age, y = count)) +
  geom_histogram(aes(y = ..count.., fill = sex), color = 'white', bins = 20) +
  facet_wrap(~ sex + death, nrow = 2, scales = 'free')  +
  labs(title = "Patient Death Status by Age and Gender",
       x = "Age",
       y = "Amount of Patients") +
  theme(legend.position = 'none') 
```




Exploratory Analysis Among Boolean Predictors
```{r}
#Death among those patients who are smokers
heart %>% group_by(death,smoking) %>% 
  summarise(n_patients = n(),
            avg_age = mean(age)) %>% 
  mutate(freq = n_patients / sum(n_patients))

#From this grouped dataframe we can see how smoking has affected deaths among patients
#Among the 299 patient, 96 were identified as smokers (32%)
#Among the 96 smokers, 30 died (31%)
```
```{r}
#Death among those patients with anaemia
heart %>% group_by(death,anaemia) %>% 
  summarise(n_patients = n(),
            avg_age = mean(age)) %>% 
  mutate(freq = n_patients / sum(n_patients))

#From this grouped dataframe we can see how anameia has affected deaths among patients
#Among the 299 patients, 129 were identified as having anaemia (43%)
#Among the 129 patients with anaemia, 46 died (35%)
```
```{r}
#Death among those patients with diabetes
heart %>% group_by(death,diabetes) %>% 
  summarise(n_patients = n(),
            avg_age = mean(age)) %>% 
  mutate(freq = n_patients / sum(n_patients))

#From this grouped dataframe we can see how diabetes has affected deaths among patients
#Among the 299 patients, 125 were identified as having diabetes (41%)
#Among the 125 patients with diabetes, 40 died (32%)
```
```{r}
#Death among those patients with high_blood_pressure
heart %>% group_by(death,high_blood_pressure) %>% 
  summarise(n_patients = n(),
            avg_age = mean(age)) %>% 
  mutate(freq = n_patients / sum(n_patients))

#From this grouped dataframe we can see how diabetes has affected deaths among patients
#Among the 299 patients, 105 were identified as having high blood pressure (35%)
#Among the 105 patients with high blood pressure, 39 died (37%)
```
```{r, fig.height= 6, fig.width=11 fig.cap = "Red = Male and Blue = Female}
#Visulization for 4 Boolean predictors
bool1 <- ggplot(data = heart,aes(x = smoking, y = count)) +
  geom_bar(aes(y = ..count.., fill = death), color = 'white', bins = 20, position = 'dodge') + 
  theme(legend.position = 'none') +
  labs(x = 'Smoking', y = '')

bool2 <- ggplot(data = heart,aes(x = anaemia, y = count)) +
  geom_bar(aes(y = ..count.., fill = death), color = 'white', bins = 20, position = 'dodge') + 
  theme(legend.position = 'none') +
    labs(x = 'Anaemia', y = '')

bool3 <- ggplot(data = heart,aes(x = diabetes, y = count)) +
  geom_bar(aes(y = ..count.., fill = death), color = 'white', bins = 20, position = 'dodge') + 
  theme(legend.position = 'none') +
    labs(x = 'Diabetes', y = '')

bool4 <- ggplot(data = heart,aes(x = high_blood_pressure, y = count)) +
  geom_bar(aes(y = ..count.., fill = death), color = 'white', bins = 20, position = 'dodge') + 
  theme(legend.position = 'none') +
    labs(x = 'High Blood Pressure', y = '')
grid.arrange(bool1,bool2,bool3,bool4,nrow = 2, top = "Count of Death for 4 Variables, Grouped by Gender")
```



Exploratory Analysis Among Numerical Predictors
```{r}
#Creatinine Phosphokinase (CPK), level of CPK in blood, Measured in Miligrams per litre (mcg/L)
heart %>% group_by(death) %>% 
  summarise(n_patients = n(),
             min_creatinine_ph = min(creatinine_ph),
             avg_creatinine_ph = mean(creatinine_ph),
             max_creatinine_ph = max(creatinine_ph),
             st_dev = sd(creatinine_ph))

#This grouped data frame gives more insight into levels of CPK in patients
#CPK on average was 130.1434 mcg/L higher for those who died
#CPK standard deviation is higher for those who dies as well
```
```{r}
#Ejection Fraction, percent of blood leaving the heart at each contraction, percentage 
heart %>% group_by(death) %>% 
  summarise(n_patients = n(),
             min_ejection = min(ejection_fraction),
             avg_ejection = mean(ejection_fraction),
             max_ejection = max(ejection_fraction),
             st_dev = sd(ejection_fraction))

#This grouped data frame gives more insight into ejection fraction among patients 
#Ejection Fraction on average was 6.79% lower for those who died
#Max ejection was 10% lower for those who died
```
```{r}
#Platelets, platelets in blood, measured in kiloplatelets per milolitre (k/ml)
heart %>% group_by(death) %>% 
  summarise(n_patients = n(),
             min_plate = min(platelets),
             avg_plate = mean(platelets),
             max_plate = max(platelets),
             st_dev = sd(platelets))

#This grouped data frame gives more insight into platelets among patients 
#Platelets on average were 10,276 k/ml higher for those who did not die
```
```{r}
#Serum Creatinine, level of of serum creatinine in blood, measured in miligrams per decilitre (mg/dl)
heart %>% group_by(death) %>% 
  summarise(n_patients = n(),
             min_sc = min(creatinine_se),
             avg_sc = mean(creatinine_se),
             max_sc = max(creatinine_se),
             st_dev = sd(creatinine_se))

#This grouped data frame gives more insight into Serum Creatinine in patients 
#Serum Creatinine on average was 0.65 mg/dl higher for those who died
```
```{r}
#Serum Sodiun, level of Serum Sodium in the blood, measured in milliequivalents per litre (mEq/L)
heart %>% group_by(death) %>% 
  summarise(n_patients = n(),
             min_ss = min(serum_sodium),
             avg_ss = mean(serum_sodium),
             max_ss = max(serum_sodium),
             st_ss = sd(serum_sodium))

#This grouped data frame gives more insight into Serum Sodium in patients 
#Serum Sodium on average was 1.84 mEq/L higher for those who did not die
```
```{r}
#correlation between numerical predictors
vars <- select(model_df,death,creatinine_se,ejection_fraction,platelets,creatinine_ph,serum_sodium)
corrplot(cor(vars),type = 'upper', method = 'color', diag = FALSE,
         tl.srt = 45, addCoef.col = 'black',
         main = 'Correlation Between Numerial Predictors and Death')
```
```{r}
#correlation between all variables
model.matrix(~death+., data = model_df) %>% cor(use = 'pairwise.complete.obs') %>%
  ggcorrplot(show.diag = F, type = 'lower', lab = TRUE, lab_size = 2, title = 'Correlation Between Each variable and Death')
```



Change dataframe factors for model
```{r}
#read in dataset again, factors were changed for visulization in exploratory analysis
model_df <- read.table(file.choose(), sep = ",", header = TRUE)
model_df <- rename(model_df, creatinine_ph = creatinine_phosphokinase, creatinine_se = serum_creatinine, death = DEATH_EVENT)
```
```{r}
#factor for model 
model_df$death <- factor(model_df$death)
model_df$anaemia <- factor(model_df$anaemia)
model_df$diabetes <- factor(model_df$diabetes)
model_df$high_blood_pressure <- factor(model_df$high_blood_pressure)
model_df$sex <- factor(model_df$sex)
model_df$smoking <- factor(model_df$smoking)
str(model_df)
```




#research questions and models 

1. "How do certain variables effect death?  Do some variables have a higher effect on death as an outcome than others?"

2. "Can a model be created to predict whether a patient will die or not based on the avaliable variables?"




Model #1, Analysis and Visulization

model 1 will attempt predict whether a paitent will die or not. This will be done with a logistic model. The response variable will be death. 


#logistic
```{r}
#split
set.seed(202)
logistic_train = sample(2,nrow(model_df), replace = T, prob = c(0.5,0.5))
train <- model_df[logistic_train == 1,]
test <- model_df[logistic_train == 2,]
#model 
glm_model <- glm(death ~ time + smoking + anaemia + diabetes + high_blood_pressure + sex + platelets + creatinine_ph + creatinine_se + ejection_fraction + serum_sodium + age, data = train, family = 'binomial')
summary(glm_model)
```
```{r}
#train prediction 
train_predict <- predict(glm_model,train, type = 'response')
print(train_predict)
print(head(train))
```
```{r}
#table and error
table <- ifelse(train_predict > 0.5,1,0)
log_table <- table(table,train$death)
print(log_table)
1 - sum(diag(log_table))/sum(log_table)
```
```{r}
#test prediction
test_predict <- predict(glm_model, test, type = 'response')
table2 <- ifelse(test_predict > 0.5,1,0)
log_table_2 <- table(table2,test$death)
print(log_table_2)
1 - sum(diag(log_table_2))/sum(log_table_2)
```
```{r}
#test model of significance 
with(glm_model, pchisq(null.deviance - deviance, df.null-df.residual, lower.tail = FALSE))
```

```{r}
glm()
```

#model 1 reran, with dropping p-values greater than .10
```{r}
#split
set.seed(203)
logistic_train2 = sample(2,nrow(model_df), replace = T, prob = c(0.5,0.5))
train2 <- model_df[logistic_train2 == 1,]
test2 <- model_df[logistic_train2 == 2,]
#model 
glm_model2 <- glm(death ~ time + creatinine_se + ejection_fraction + age, data = train2, family = 'binomial')
summary(glm_model2)
```
```{r}
#train prediction 
train_predict2 <- predict(glm_model2,train2, type = 'response')
print(train_predict2)
print(head(train2))
```
```{r}
#table and error
table2 <- ifelse(train_predict2 > 0.5,1,0)
log_table2 <- table(table2,train2$death)
print(log_table2)
1 - sum(diag(log_table2))/sum(log_table2)
```
```{r}
#test prediction
test_predict2 <- predict(glm_model2, test2, type = 'response')
table3 <- ifelse(test_predict2 > 0.5,1,0)
log_table_3 <- table(table3,test2$death)
print(log_table_3)
1 - sum(diag(log_table_3))/sum(log_table_3)
```
```{r}
#test model of significance 
with(glm_model, pchisq(null.deviance - deviance, df.null-df.residual, lower.tail = FALSE))
```








Model #2, Analysis and Visulization
 
To awnser this question, a classification tree will executed and analyzed. Time will be dropped from model, time is not useful for the model. Time variable is represented as the last time the patient had a check up, if they are dead they did not have one. Time would have a much higher effect for taying alive. for this we are trying to find the effects on death. 

```{r}
#fitting classification tree, small cp before trim
set.seed(2)
rpart.model_df = rpart(death ~ time + age + anaemia + creatinine_ph + diabetes + ejection_fraction + high_blood_pressure + platelets + creatinine_se + serum_sodium + sex + smoking,
                       data = heart,
                       cp = 0.00000000001)
```
```{r}
#model results
plotcp(rpart.model_df)
printcp(rpart.model_df)
```
Min xerror = .84375, associated xstd = .080050, associtated cp = .01

```{r}
#prune
rpart.model_df.min <- prune(rpart.model_df, cp = 0.01)
rpart.plot(rpart.model_df.min, extra = 2, digits = 3, main = 'Death Classification Tree')
```
```{r}
#compare cross-validated full-set error with test set error
set.seed(2)
train_2 = sample(1:nrow(heart), size = 149)
test_2 = heart[-train_2,]
death.test = heart$death[-train_2]
#rerun with apporpriate cp and train
set.seed(2)
rpart.train_2 = rpart(death ~ time + age + anaemia + creatinine_ph + diabetes + ejection_fraction + high_blood_pressure + platelets + creatinine_se + serum_sodium + sex + smoking,
                       data = heart, subset = train_2, cp = 0.01)
#prediction class type
rpart.pred = predict(rpart.train_2, test_2, type = 'class')
table(rpart.pred, death.test)
#mean
mean(rpart.pred!=death.test)
```
#bagged trees for importance
```{r}
#model 
train = sample(1:nrow(model_df), nrow(model_df)/2)
set.seed(9301)
bag.model = randomForest(death ~ time + age + anaemia + creatinine_ph + diabetes + ejection_fraction + high_blood_pressure + platelets + creatinine_se + serum_sodium + sex + smoking , data = heart, subset = train, mtry =11, importance = TRUE)
```
```{r}
bag.model
```
```{r}
importance(bag.model)
```












